---
title: "Academic Impact Part 2"
subtitle: "Data Visualization and Analysis in the Era of COVID-19"
author: "Damian Betebenner, Adam VanIwaarden & Nathan Dadey"
institute: "Center for Assessment"
date: "March 22nd, 2022 (updated: `r toOrdinal::toOrdinalDate(Sys.Date())`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: ["center", "middle", "my-title"]

---
class: inverse, center, middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r share-again, echo=FALSE}
xaringanExtra::use_share_again()
```

```{r style-share-again, echo=FALSE}
xaringanExtra::style_share_again(
  share_buttons = c("twitter"),
  background = "#009C49",
  foreground = "#FFFFFF"
)
```

```{r xaringan-logo, echo=FALSE}
xaringanExtra::use_logo(
  image_url = "https://centerforassessment.github.io/assets/favicon.png",
  width = "55px",
  height = "64px",
  position = xaringanExtra::css_position(top = "0.5em", right = "0.5em"),
  link_url = "https://www.nciea.org"
)
```

```{r xaringan-tile-view, echo=FALSE}
xaringanExtra::use_tile_view()
```

```{r xaringanExtra-clipboard, echo=FALSE}
xaringanExtra::use_clipboard()
```

```{r load-packages, include=FALSE}
library(data.table)
library(SGP)
library(SGPdata)
library(splines)
```

# Academic Impact

---

## Learning Deceleration/Acceleration

--

- Summarization of assessment data (either for diagnostic or accountability purposes) of emphasizes two types of data results: Status (i.e. student attainment) and Growth (i.e., student academic progress).

- With the pandemic, emaphasis has changed to look at _academic impact_.

- In yesterday's introduction we discussed status impact (odometer) and growth impact (speedometer).

- We emphasize that Impact is synonymous with _decleration_ and Recovery is synonymous with _acceleration_.

- Deceleration, by definition, is the change in velocity (i.e. growth).

- Improvement (increasing student attainment/status in a system over time) is indicative of acceleration (unless the population is changing leading to the increased attainment/status)

- One of the most informative and least appreciated papers that discusses this is [Focusing State Educational Accountability Systems: Four Methods of Judging School Quality and Progress](https://www.nciea.org/sites/default/files/publications/Dale020402.pdf)


---

```{r echo=FALSE}
magick::image_read_pdf("Figures/Carlson_Four_Quadrant_Figure.pdf")
```

---

# Data to Investigate Academic Impact

- The data set we will utilize is called __sgpData_LONG_COVID__ and is a part of the __SGPdata__ package you've installed.

- Load the package into your R environment:


- Explore the dataset:

```{r eval=FALSE, tidy=FALSE}
summary(sgpData_LONG_COVID)
```

---

# Status Based Academic Impact

- To analyze and visualize status based impact we will investigate changes to the Empirical Cummulative Distribution Functions (ECDFs) for grade by cohort pairs.

- We begin by analyzing ECDFs for Grade 8 mathematics in 2019 versus Grade 8 mathematics in 2021.

```{r eval=TRUE, tidy=FALSE, cache=TRUE}
math_g8_2019 <- sgpData_LONG_COVID[CONTENT_AREA=="MATHEMATICS" & GRADE=="8" & YEAR=="2019"]
math_g8_2021 <- sgpData_LONG_COVID[CONTENT_AREA=="MATHEMATICS" & GRADE=="8" & YEAR=="2021"]
math_g8_2019_quantiles <- quantile(math_g8_2019$SCALE_SCORE, probs=0:100/100)
math_g8_2021_quantiles <- quantile(math_g8_2021$SCALE_SCORE, probs=0:100/100)
```

---

## Academic Impact: Status

```{r echo=FALSE}
magick::image_read_pdf("Figures/Academic_Impact_Status_GRID.pdf")
```

---

## Create Plot of Status Based Academic Impact

```{r eval=TRUE, tidy=FALSE, fig.height=4, fig.align='center'}
par(mai=c(0.5, 0.5, 0.25, 0))
plot(x=50, y=500, type="n", xlim=c(0,100), ylim=c(350, 700), axes=FALSE, main="Status Academic Impact: Mathematics Grade 8")
points(0:100, math_g8_2019_quantiles, type="l", col="black")
points(0:100, math_g8_2021_quantiles, type="l", col="red")
axis(1, at=0:10*10)
axis(2, at=c(350, 400, 450, 500, 550, 600, 650, 700), cex.axis=0.7)
```

---

## Create Plot of Status Based Academic Impact

```{r eval=TRUE, tidy=FALSE, fig.height=4, fig.align='center'}
par(mai=c(0.5, 0.5, 0.25, 0))
plot(x=50, y=500, type="n", xlim=c(0,100), ylim=c(350, 700), axes=FALSE, main="Status Academic Impact: Mathematics Grade 8")
points(0:100, math_g8_2019_quantiles, type="l", col="black")
points(0:100, math_g8_2021_quantiles, type="l", col="red")
axis(1, at=0:10*10)
axis(2, at=c(350, 400, 450, 500, 550, 600, 650, 700), cex.axis=0.7)
abline(h=c(472, 500, 542), lty=2, col="blue")
```

---

## Identifying percent at/above proficient

- Normally, you will have the percentage of students in a grade/content area at above proficient.
- However, they can be derived from the data we are using.
- From above, we know that 577 is the proficient cutscore for grade 8, mathematics
- We can use ECDFs to calculate these

```{r eval=TRUE, cache=TRUE}
math_g8_2019_ecdf <- ecdf(math_g8_2019$SCALE_SCORE)
math_g8_2021_ecdf <- ecdf(math_g8_2021$SCALE_SCORE)
100 - 100*math_g8_2019_ecdf(500)
100 - 100*math_g8_2021_ecdf(500)
```

---

```{r eval=TRUE, tidy=FALSE, fig.height=4, fig.align='center'}
par(mai=c(0.5, 0.5, 0.25, 0))
plot(x=50, y=500, type="n", xlim=c(0,100), ylim=c(350, 700), axes=FALSE, main="Status Academic Impact: Mathematics Grade 8")
points(0:100, math_g8_2019_quantiles, type="l", col="black")
points(0:100, math_g8_2021_quantiles, type="l", col="red")
axis(1, at=0:10*10)
axis(2, at=c(350, 400, 450, 500, 550, 600, 650, 700), cex.axis=0.7)
abline(h=c(472, 500, 542), lty=2, col="blue")
abline(v=c(55.26, 43.58), col="black", lty=3)
```

---

## To facilitate plotting convert LONG data to WIDE data

```{r eval=FALSE}
SGP::outputSGP(sgpData_LONG_COVID, output.type='WIDE')
```

---

## Growth Based Academic Impact

```{r eval=TRUE, tidy=FALSE, fig.height=4, fig.align='center'}
load('Data/Demonstration_COVID_SGP_WIDE_Data.Rdata')
g8_2019_2021_data <- Demonstration_COVID_SGP_WIDE_Data[GRADE.2019.MATHEMATICS=="6" & GRADE.2021.MATHEMATICS=="8"]
g8_2017_2019_data <- Demonstration_COVID_SGP_WIDE_Data[GRADE.2017.MATHEMATICS=="6" & GRADE.2019.MATHEMATICS=="8"]
attach(g8_2019_2021_data)
plot(SCALE_SCORE.2019.MATHEMATICS, SCALE_SCORE.2021.MATHEMATICS, pch=".", cex=0.5, xlab="Scaled Score 2019: Grade 6", ylab="Scaled Score 2021: Grade 8")
detach(g8_2019_2021_data)
```

---

## Calculate conditional status splines

```{r eval=FALSE}
load('Data/Demonstration_COVID_SGP_WIDE_Data.Rdata')
g8_2019_2021_data <- Demonstration_COVID_SGP_WIDE_Data[GRADE.2019.MATHEMATICS=="6" & GRADE.2021.MATHEMATICS=="8"]
g8_2017_2019_data <- Demonstration_COVID_SGP_WIDE_Data[GRADE.2017.MATHEMATICS=="6" & GRADE.2019.MATHEMATICS=="8"]
attach(g8_2019_2021_data)
spline_2019_to_2021 <- lm(SCALE_SCORE.2021.MATHEMATICS ~ bs(SCALE_SCORE.2019.MATHEMATICS))
detach(g8_2019_2021_data)

attach(g8_2017_2019_data)
spline_2017_to_2019 <- lm(SCALE_SCORE.2019.MATHEMATICS ~ bs(SCALE_SCORE.2017.MATHEMATICS))
detach(g8_2017_2019_data)
```

---


```{r eval=FALSE}
attach(g8_2019_2021_data)
plot(SCALE_SCORE.2019.MATHEMATICS, SCALE_SCORE.2021.MATHEMATICS, pch=".", cex=0.5, xlab="Scaled Score 2019: Grade 6", ylab="Scaled Score 2021: Grade 8")
x_coors <- min(SCALE_SCORE.2019.MATHEMATICS, na.rm=TRUE):max(SCALE_SCORE.2019.MATHEMATICS)
points(x_coors, predict(spline_2017_to_2019, data.frame(SCALE_SCORE.2017.MATHEMATICS=x_coors)), col="black", type="l")
points(x_coors, predict(spline_2019_to_2021, data.frame(SCALE_SCORE.2019.MATHEMATICS=x_coors)), col="red", type="l")
detach(g8_2019_2021_data)
```

---

```{r echo=FALSE}
magick::image_read_pdf("Figures/Grade_6_2019_to_Grade_8_2021_Math.pdf")
```
---
