---
title: "Data Visualization and Analysis in the Era of COVID-19"
subtitle: "Software Preparations"
author: "Damian Betebenner, Adam VanIwaarden & Nathan Dadey"
institute: "Center for Assessment"
date: "March 3rd, 2022 (updated: `r toOrdinal::toOrdinalDate(Sys.Date())`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: ["center", "middle", "my-title"]

---
class: inverse, center, middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r share-again, echo=FALSE}
xaringanExtra::use_share_again()
```

```{r style-share-again, echo=FALSE}
xaringanExtra::style_share_again(
  share_buttons = c("twitter")
)
```

```{r xaringan-logo, echo=FALSE}
xaringanExtra::use_logo(
  image_url = "https://centerforassessment.github.io/assets/favicon.png"
)
```

```{r xaringan-tile-view, echo=FALSE}
xaringanExtra::use_tile_view()
```

```{r xaringanExtra-clipboard, echo=FALSE}
xaringanExtra::use_clipboard()
```

# Required Software

---

# R

--

## We do all of our data analysis using the [R Software Environment](https://cran.r-project.org/)

--

- We assume that you all have some familiarity with using R.

--

- If you don't already have R installed, please go ahead and install it from [CRAN](https://cran.r-project.org/) picking the version that goes with your operating systems.

- If you have R installed but have an older version (pre 4.1), now is a good time to update the version that you're using.

- Many of you familiar with R probably use the [RStudio development environment](https://www.rstudio.com/) instead of the default user interface provided with R.


---

# R Packages

## One of R's greatest strengths is its extensibility.

--

- One can create their own R packages that contain custom data and code to do just about anything.

--

- Over the last 15 years, we have done growth analyses for more than two dozen states using R and the [SGP Package](https://www.sgp.io) for calculating student growth percentiles and percentile growth trajectories.

--

- Packages can be distributed via the [Central R Archive Network](https://cran.r-project.org/web/packages/available_packages_by_name.html) (CRAN). As of today there are 19,014 packages on CRAN.

--

- Many amazing R packages are currently being built and distributed using [GitHub](www.github.com). Developing and allowing others to install packages from GitHub is by far the best way to do serious data analysis with R.

--

- This presentation was created using the R package [Xaringan](https://bookdown.org/yihui/rmarkdown/xaringan.html)

--

- Have any of you created an R package?

---

# Required R packages for this training session

## We will need a few packages for this training session.

--

- [SGP Package on CRAN](https://cran.rstudio.com/web/packages/SGP/index.html) or the more current [SGP Package on GitHub](https://github.com/CenterForAssessment/SGP)

- [SGPdata Package on CRAN](https://cran.rstudio.com/web/packages/SGPdata/index.html) or the more updated [SGPdata Package on GitHub](https://github.com/CenterForAssessment/SGPdata)

- [cfaTools package on GitHub](https://centerforassessment.github.io/cfaTools/) (There isn't a CRAN version of this package yet.)

---

# Let's install the SGP package

### To install the SGP package from CRAN and GitHub use the following commands (you can copy and past them into your R development environement of choice)

### RStudio has several pull down menu options in its IDE that (essentially) runs the following commands for you.

### CRAN

```{r eval=FALSE, tidy=FALSE}
install.packages("SGP")
```

### GitHub

```{r eval=FALSE, tidy=FALSE}
remotes::install_github("centerforassessment/SGP")
```

---

# Let's install the SGPdata package

### CRAN

```{r eval=FALSE, tidy=FALSE}
install.packages("SGPdata")
```

### GitHub

```{r eval=FALSE, tidy=FALSE}
remotes::install_github("centerforassessment/SGPdata")
```

---

# Let's install the cfaTools package

### Not available on CRAN

### GitHub

```{r eval=FALSE, tidy=FALSE}
remotes::install_github("centerforassessment/cfaTools")
```

---

# Let's install the rest of the packages

```{r eval=FALSE, tidy=FALSE}
install.packages(c("data.table", "VIM", "mice", "miceadds", "ggplot2", "lme4", "merTools", "gghighlight"))
```
