---
title: "Data Visualization and Analysis in the Era of COVID-19"
subtitle: "Software Preparations"
author: "Damian Betebenner, Adam VanIwaarden & Nathan Dadey"
institute: "Center for Assessment"
date: "March 3rd, 2022 (updated: `r toOrdinal::toOrdinalDate(Sys.Date())`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: ["center", "middle", "my-title"]

---
class: inverse, center, middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r share-again, echo=FALSE}
xaringanExtra::use_share_again()
```

```{r style-share-again, echo=FALSE}
xaringanExtra::style_share_again(
  share_buttons = c("twitter")
)
```

```{r xaringan-logo, echo=FALSE}
xaringanExtra::use_logo(
  image_url = "https://centerforassessment.github.io/assets/favicon.png"
)
```

```{r xaringan-tile-view, echo=FALSE}
xaringanExtra::use_tile_view()
```

```{r xaringanExtra-clipboard, echo=FALSE}
xaringanExtra::use_clipboard()
```

# Required Software

---

# R

--

## We do all of our data analysis using the [R Software Environment](https://cran.r-project.org/)

--

- We assume that you all have some familiarity with using R.

--

- If you don't already have R installed, please go ahead an install it from [CRAN](https://cran.r-project.org/)

--


---

# R Packages

## One of R's greatest strengths is its extensibility.

--

- One can create their own R packages that contain custom code to do just about anything.

--

- Over the last 15 years, we have done growth analyses for more than two dozen states using R and the [SGP Package](https://www.sgp.io) for calculating student growth percentiles and percentile growth trajectories.

--

- Packages can be distributed via the [Central R Archive Network (CRAN)](https://cran.r-project.org/web/packages/available_packages_by_name.html). As of today there are 19,014 packages on CRAN.

--

- Many amazing R packages are currently being built and distributed using [GitHub](www.github.com). Developing and allowing others to install packages from GitHub is by far the best way to do serious data analysis with R.

--

- This presentation was created using the R package [Xaringan](https://bookdown.org/yihui/rmarkdown/xaringan.html)

--

- Have any of you created and R package?

---

# Required R packages for this training session

## We will need a few packages for this training session.

--

- [SGP Package CRAN](https://cran.rstudio.com/web/packages/SGP/index.html)

- [SGP Package GitHub](https://github.com/CenterForAssessment/SGP)

- [SGPdata Package CRAN](https://cran.rstudio.com/web/packages/SGPdata/index.html)

- [SGPdata Package GitHub](https://github.com/CenterForAssessment/SGPdata)

---

# Let's install the required packages
